---
title: "Workflow: from a Stacks run to a results table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow: from a Stacks run to a results table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through one analysis from start to finish, on the toy
dataset that ships with the package: 80 RAD loci and two populations of 4
and 3 individuals. That is far too few individuals for real inference -- it
is here so every step runs in seconds. The reasoning behind each step is in
`vignette("rationale", package = "RADdiversity")`.

```{r}
library(RADdiversity)
haps   <- system.file("extdata", "small.haps.vcf", package = "RADdiversity")
snps   <- system.file("extdata", "small.snps.vcf", package = "RADdiversity")
popmap <- system.file("extdata", "small_popmap.tsv", package = "RADdiversity")
sumstats <- system.file("extdata", "populations.sumstats_summary.tsv",
                        package = "RADdiversity")
```

## 1. Run Stacks once

A few `populations` flags are worth setting carefully for everything below: a
minimum genotype depth, so that heterozygotes are not silently called
homozygous, and a per-population (not pooled) missing-data filter, so that
both populations are filtered the same way. There is deliberately no `--hwe`
filter.

```bash
populations --in-path ./stacks_out --popmap popmap.tsv -O ./out \
            --min-gt-depth 6 -r 0.8 -p 2 --min-mac 3 \
            --max-obs-het 0.70 --fstats --vcf --genepop --fasta-samples -t 8
```

This writes `populations.snps.vcf` (one record per SNP) and
`populations.haps.vcf` (one record per RAD locus), and
`populations.sumstats_summary.tsv`. Add `--vcf-all` to also get every
sequenced site, variable or not, for `pi_allsites()` (section 8).

## 2. Read the data

```{r}
H <- read_stacks_vcf(haps)
pops <- read_popmap(popmap, H$samples)
```

`read_stacks_vcf()` also reads VCFs from other pipelines (ipyrad, dDocent,
GATK ...). What matters is that each record is assigned to its RAD locus --
the unit that standard errors resample -- and the message above says how that
was done; see `locus_from` in `?read_stacks_vcf`.

## 3. Optional: filter loci

The filters take and return the same object, so they chain. Looking before
filtering is cheap:

```{r}
head(locus_allele_stats(H))
H_f <- filter_call_rate(H, min_call = 0.5, pops = pops)
H_f <- filter_max_het(H_f, max_ho = 0.8)
```

`filter_maf()`, `filter_mac()`, `filter_thin_one_snp()` and
`filter_low_conf_alt()` work the same way. A filtered object can go straight
into the functions below (with `stem =` naming its output files).

## 4. Screen for close relatives

Every between-population test below assumes independent individuals.

```{r}
kin <- kinship_check(H)
kin$flagged_pairs
```

KING-robust misses many real relatives in reduced-representation data
(McMaster et al. 2025), so a clean result is reassuring, not proof.

## 5. Diversity within each population

Run `diversity_stats()` on both files: the haplotype VCF gives FIS and
allelic richness, the SNP VCF gives Ho, He and their per-site values.

```{r}
div_haps <- diversity_stats(haps, popmap, g = 4, nboot = 200)
div_haps$per_population
div_haps$richness
```

```{r}
div_snps <- diversity_stats(snps, popmap, g = 4, nboot = 200, sites = sumstats)
div_snps$autosomal
```

Printing a result shows the full report, including which of its numbers to
take from this file:

```{r, eval = FALSE}
print(div_snps)
```

## 6. Do individuals differ in inbreeding?

The standard errors and intervals above resample loci, holding the sampled
individuals fixed. That is fine when individuals are exchangeable and too
optimistic when they differ in inbreeding. g2 answers the question:

```{r}
identity_disequilibrium(H, pops, nboot = 200, nperm = 200, min_call = 0.5)
```

If its interval excludes 0, report the individual-jackknife standard errors:

```{r}
div_ind <- diversity_stats(haps, popmap, g = 4, nboot = 0, se_individuals = TRUE)
div_ind$per_population[, c("population", "Fis", "Fis_se", "Fis_se_ind")]
```

## 7. Do the populations differ?

Compare individuals, not loci. `het_between_pops()` tests two things with the
same machinery: mean individual heterozygosity (diversity) and the individual
inbreeding coefficient F (inbreeding):

```{r}
het <- het_between_pops(haps, popmap, min_call = 0.5)
het$pairwise_tests      # heterozygosity
het$pairwise_F_tests    # inbreeding
```

Differentiation between the populations:

```{r}
dif <- differentiation_stats(haps, popmap, nboot = 200)
dif$global
```

## 8. Nucleotide diversity from an all-sites VCF

With `populations --vcf-all` output, `pi_allsites()` computes pi and dxy
directly, handling missing data site by site. A tiny file shows the shape:

```{r}
allsites <- tempfile(fileext = ".vcf")
writeLines(c(
  "##fileformat=VCFv4.2",
  "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\ta1\ta2\tb1\tb2",
  "1\t1\t.\tA\tG\t.\tPASS\t.\tGT\t0/0\t0/1\t1/1\t1/1",
  "1\t2\t.\tC\t.\t.\tPASS\t.\tGT\t0/0\t0/0\t0/0\t0/0",
  "2\t1\t.\tT\tC\t.\tPASS\t.\tGT\t0/1\t./.\t0/0\t0/1"), allsites)
pm4 <- tempfile()
writeLines(c("a1\tpopA", "a2\tpopA", "b1\tpopB", "b2\tpopB"), pm4)
pi_allsites(allsites, pm4, nboot = 0)$pi
```

## 9. Hardy-Weinberg: a report, never a filter

A heterozygote deficit is the signal FIS measures, so no function here removes
loci on HWE:

```{r}
hw <- hwe_test(H, pops = pops, n_draws = 1000)
table(p_below_0.05 = hw$p_value < 0.05, useNA = "ifany")
```

## 10. Export

```{r, eval = FALSE}
write_vcf(H_f, "filtered.vcf")
write_plink(H_f, "filtered", pops = pops, drop_multiallelic = TRUE)
write_structure(H_f, "filtered.str", pops = pops)
write_genepop(H_f, "filtered.gen", pops = pops)
write_radpainter(H_f, "filtered.radpainter")
```

## What to write

Methods and results text to adapt, and answers to the usual reviewer
objections, are in `vignette("reviewer-faq", package = "RADdiversity")`.
