---
title: "Workflow: from a Stacks run to a results table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow: from a Stacks run to a results table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This vignette walks through one analysis from start to finish, on the toy
dataset that ships with the package: 80 RAD loci and two populations of 4
and 3 individuals. That is far too few individuals for real inference -- it
is here so every step runs in seconds. The reasoning behind each step is in
`vignette("rationale", package = "RADdiversity")`.

```{r}
library(RADdiversity)
haps     <- system.file("extdata", "small.haps.vcf", package = "RADdiversity")
snps     <- system.file("extdata", "small.snps.vcf", package = "RADdiversity")
popmap   <- system.file("extdata", "small_popmap.tsv", package = "RADdiversity")
sumstats <- system.file("extdata", "populations.sumstats_summary.tsv",
                        package = "RADdiversity")

# Bootstrap intervals and permutation p-values use random numbers. Setting the
# seed once makes every result below reproducible.
set.seed(2024)
```

Two conventions hold for every analysis function:

* the data (`vcf`) can be a file path or an object you have already read
  with `read_stacks_vcf()` (and perhaps filtered);
* the populations (`popmap`) can be a file path or the list returned by
  `read_popmap()`.

Printing a result shows its main tables; `summary()` shows the full report
with notes on interpretation. Every table is also an element of the result
(`res$per_population`, ...), stored at full precision. Add `verbose = FALSE`
to any call to silence its progress messages.

## 1. Run Stacks once

A few `populations` flags are worth setting carefully for everything below: a
minimum genotype depth, so that heterozygotes are not silently called
homozygous, and a per-population (not pooled) missing-data filter, so that
both populations are filtered the same way. There is deliberately no `--hwe`
filter, and no `--min-mac`/`--min-maf`: Stacks turns a SNP that fails them
into a fixed site, and the rare variants they remove carry much of the
diversity (about 10% of π at 20 diploids, 21% at 10). Make a separate,
filtered run if another analysis needs one.

```bash
populations --in-path ./stacks_out --popmap popmap.tsv -O ./out \
            --min-gt-depth 6 -r 0.8 -p 2 \
            --max-obs-het 0.70 --fstats --vcf --genepop --fasta-samples -t 8
```

This writes `populations.snps.vcf` (one record per SNP),
`populations.haps.vcf` (one record per RAD locus), and
`populations.sumstats_summary.tsv`. Add `--vcf-all` to also get every
sequenced site, variable or not, for `pi_allsites()` (section 8).

## 2. Read the data

```{r}
H <- read_stacks_vcf(haps)
H
pops <- read_popmap(popmap, H$samples)
pops
```

`read_stacks_vcf()` also reads VCFs from other pipelines (ipyrad, dDocent,
GATK ...). What matters is that each record is assigned to its RAD locus --
the unit that standard errors resample -- and the message above says how that
was done; see `locus_from` in `?read_stacks_vcf`.

## 3. Optional: filter loci

The filters take and return the same object, so they chain. Looking before
filtering is cheap:

```{r}
head(locus_allele_stats(H))
H_f <- filter_call_rate(H, min_call = 0.5, popmap = pops)
H_f <- filter_max_het(H_f, max_ho = 0.8)
```

`filter_genotype_depth()`, `filter_maf()`, `filter_mac()`,
`filter_thin_one_snp()` and `filter_low_conf_alt()` work the same way. A
filtered object can go straight into the functions below (add `stem = ` to
name its output files if you use `outdir`). Two cautions for diversity
statistics: an allele-frequency filter (`filter_maf()`, `filter_mac()`)
removes low-diversity sites and biases H<sub>e</sub> per site, and
`filter_low_conf_alt()` masks only calls carrying an ALT allele, which
removes heterozygotes preferentially; to mask low-depth genotypes use
`filter_genotype_depth()`, which treats every genotype alike.

## 4. Screen for close relatives

Every between-population test below assumes independent individuals.

```{r}
kin <- kinship_check(H, pops)
kin$flagged_pairs
```

KING-robust misses many real relatives in reduced-representation data
(McMaster et al. 2025), so a clean result is reassuring, not proof. With
`method = "beta"`, always pass the popmap: beta compares each pair with the
average pair it is given, so on several populations pooled, unrelated
members of the same population read as relatives.

## 5. Diversity within each population

Run `diversity_stats()` on both files: the haplotype VCF gives FIS and
allelic richness, the SNP VCF gives Ho, He and their per-site values.

```{r}
div_haps <- diversity_stats(haps, pops, g = 4, nboot = 200)
div_haps
```

```{r}
div_snps <- diversity_stats(snps, pops, g = 4, nboot = 200, sites = sumstats,
                            verbose = FALSE)
div_snps$autosomal
```

The full report explains each number, including which of them to take from
this file:

```{r, eval = FALSE}
summary(div_snps)
```

## 6. Do individuals differ in inbreeding?

The standard errors and intervals above resample loci, holding the sampled
individuals fixed. That is fine when individuals are exchangeable and too
optimistic when they differ in inbreeding. g2 answers the question:

```{r}
identity_disequilibrium(H, pops, nboot = 200, nperm = 200, min_call = 0.5)
```

If its interval excludes 0, report the individual-jackknife standard errors:

```{r}
div_ind <- diversity_stats(haps, pops, g = 4, nboot = 0, se_individuals = TRUE,
                           verbose = FALSE)
div_ind$per_population[, c("population", "Fis", "Fis_se", "Fis_se_ind")]
```

## 7. Do the populations differ?

Compare individuals, not loci. `het_between_pops()` tests two things with the
same machinery: mean individual heterozygosity (diversity) and the individual
inbreeding coefficient F (inbreeding):

```{r}
het <- het_between_pops(haps, pops, min_call = 0.5, verbose = FALSE)
het$pairwise_tests      # heterozygosity
het$pairwise_F_tests    # inbreeding
```

Differentiation between the populations:

```{r}
dif <- differentiation_stats(haps, pops, nboot = 200, verbose = FALSE)
dif$global
```

## 8. Nucleotide diversity from an all-sites VCF

With `populations --vcf-all` output, `pi_allsites()` computes pi and dxy
directly, handling missing data site by site. A tiny file shows the shape:

```{r}
allsites <- tempfile(fileext = ".vcf")
writeLines(c(
  "##fileformat=VCFv4.2",
  "#CHROM\tPOS\tID\tREF\tALT\tQUAL\tFILTER\tINFO\tFORMAT\ta1\ta2\tb1\tb2",
  "1\t1\t.\tA\tG\t.\tPASS\t.\tGT\t0/0\t0/1\t1/1\t1/1",
  "1\t2\t.\tC\t.\t.\tPASS\t.\tGT\t0/0\t0/0\t0/0\t0/0",
  "2\t1\t.\tT\tC\t.\tPASS\t.\tGT\t0/1\t./.\t0/0\t0/1"), allsites)
pi_allsites(allsites, list(popA = c("a1", "a2"), popB = c("b1", "b2")), nboot = 0)$pi
```

## 9. Hardy-Weinberg: a report, never a filter

A heterozygote deficit is the signal FIS measures, so no function here removes
loci on HWE:

```{r}
hw <- hwe_test(H, pops, n_draws = 1000, verbose = FALSE)
table(p_below_0.05 = hw$p_value < 0.05, useNA = "ifany")
```

## 10. Export

```{r, eval = FALSE}
write_vcf(H_f, "filtered.vcf")
write_plink(H_f, "filtered", popmap = pops, drop_multiallelic = TRUE)
write_structure(H_f, "filtered.str", popmap = pops)
write_genepop(H_f, "filtered.gen", popmap = pops)
write_radpainter(H_f, "filtered.radpainter")
```

## What to write

Methods and results text to adapt, and answers to the usual reviewer
objections, are in `vignette("reviewer-faq", package = "RADdiversity")`.
