---
title: "Workflow: from a Stacks run to a results table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Workflow: from a Stacks run to a results table}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 90)
```

This vignette walks through one analysis from start to finish. Each step
says **what** it does, **why** it is done that way, and **how to read**
the output. The longer rationale behind the choices are in
`vignette("rationale", package = "RADdiversity")`.

## The example data

The package ships a simulated dataset shaped like Stacks 2 output: two
populations, `north` (15 individuals) and `south` (10), typed at 1,000
RAD loci of 80 bp. The simulation has a few things built in for the
analysis to find: south went through a bottleneck, individuals differ in
how inbred they are, two individuals in north are full siblings, and
missing data are clustered by locus (`inst/sims/make_example_data.R`
shows exactly how the data were made).

```{r}
library(RADdiversity)
snps     <- system.file("extdata", "example.snps.vcf.gz", package = "RADdiversity")
haps     <- system.file("extdata", "example.haps.vcf.gz", package = "RADdiversity")
allsites <- system.file("extdata", "example.allsites.vcf.gz", package = "RADdiversity")
popmap   <- system.file("extdata", "example_popmap.tsv", package = "RADdiversity")
sumstats <- system.file("extdata", "example.sumstats_summary.tsv", package = "RADdiversity")

# Bootstrap intervals and permutation p-values use random numbers; setting the
# seed once makes every result below reproducible.
set.seed(2024)
```

To use your own data, replace these paths with your files. A popmap is a
plain text file with no header: one line per individual, the sample name
(as in the VCF), a tab, and the population.

```{r}
readLines(popmap, n = 3)
```

### Five ideas used throughout

- **Gene copies.** A diploid individual carries two copies of each
  locus, so 10 individuals are 20 gene copies. Rarefaction sizes (`g`)
  are in gene copies.
- **Record versus RAD locus.** Stacks writes two VCFs if multiple SNPs
  per locus are allowed by *populations* filters. In
  `populations.snps.vcf` each record (line) is one SNP, and several SNPs
  can sit on the same short RAD locus. In `populations.haps.vcf` each
  record is one RAD locus, and its alleles are haplotypes (the SNPs of
  that locus read together).
- **Ho, He and FIS.** Observed heterozygosity (Ho) is the share of
  individuals that are heterozygous. Expected heterozygosity (He, or
  gene diversity) is the chance that two gene copies drawn from the
  population differ. FIS = 1 − Ho/He measures the heterozygote deficit:
  0 under random mating, positive with inbreeding.
- **Allelic richness.** The number of alleles seen depends on how many
  individuals you sampled, so richness is *rarefied*: the expected
  number of alleles in the same number of gene copies (`g`) for every
  population.
- **Jackknife and bootstrap.** Both measure uncertainty by recomputing a
  statistic on altered versions of the data: the jackknife leaves out
  one unit at a time, the bootstrap resamples units with replacement.
  Here the unit is normally a RAD locus, because SNPs on one locus are
  inherited together.

## 1. Where the data come from

### If you are running Stacks yourself

These `populations` settings suit the analyses below:

``` bash
populations --in-path ./stacks_out --popmap popmap.tsv -O ./out \
            -p 2 -r 0.8 --min-gt-depth 6 --max-obs-het 0.70 \
            --vcf --vcf-all -t 8
```

- `-p 2 -r 0.8` keeps a locus only if at least 80% of the individuals
  *of each population* are genotyped there, in at least 2 populations.
  Setting `-p` to your number of populations is recommended. **Why:**
  the alternative, `-R 0.#`, counts all individuals pooled, so a
  well-sequenced large population can carry a poorly sequenced small
  one. The two populations then have different amounts of missing data,
  which can look like a difference in heterozygosity.
- `--min-gt-depth 6` (Stacks 2.67 or later) sets genotypes supported by
  fewer than 6 reads to missing. **Why:** a heterozygote read only *d*
  times shows just one of its alleles with probability 2\^(1−*d*): 25%
  at 3 reads, 3% at 6. Those calls look like homozygotes and inflate
  FIS.
- `--max-obs-het 0.70` removes sites where more than 70% of individuals
  are heterozygous, a crude screen for paralogs (two similar genomic
  regions assembled as one locus).
- `--vcf` writes both VCFs; `--vcf-all` (Stacks 2.62 or later) also
  writes every sequenced site, which `pi_allsites()` uses (section 9).

Two filters are left out of this command. There is no Hardy-Weinberg
filter, because a heterozygote deficit is exactly what FIS measures
(section 10). And there is no minor allele count or frequency filter
(`--min-mac`, `--min-maf`). Those are common in RAD-seq studies and fine
to use; the next section explains what they change for diversity
statistics, so you can report them accurately.

### If your data were already filtered

Often the VCF was made by someone else, or with a pipeline that filtered
more. That is normal, and the analyses still run. What matters is
knowing what each filter changes and reporting it:

| filter applied before this package | what it changes | what to do |
|------------------------|------------------------|------------------------|
| minor allele count or frequency (`populations --min-mac`, `--min-maf`; `vcftools --mac/--maf`) | removes rare alleles: lowers He per sequenced site, % polymorphic, Ar and especially privAr, by more in smaller samples; slightly raises FIS | report the threshold; compare these values only with datasets filtered the same way |
| pooled missing-data filter (`-R`, `vcftools --max-missing`) | can leave one population with more missing data than another | check `fis_by_call_rate` (section 5) and the missingness check in `het_between_pops()` (section 7); or re-filter per population with `filter_call_rate(..., popmap = )` |
| Hardy-Weinberg filter | removes loci with heterozygote deficits, pulling FIS toward 0 | report it; if possible, re-run without it for FIS |
| maximum heterozygosity (`--max-obs-het`) | removes loci with many heterozygotes (often paralogs) | usually fine; report it |
| one SNP per RAD locus (`--write-single-snp`) | fewer SNPs; He per SNP shifts differently in different populations; per-site He comes out far too low, because most SNPs are missing | use all SNPs for He and per-site values |
| any filter applied after Stacks `populations` | Stacks' `Sites` no longer matches the data if whole RAD loci were removed | give `sites` for the loci you kept (see "The SNP run" in section 5) |

`summary()` of a `diversity_stats()` result (section 5) shows what the
data look filtered at: the rarest allele count left, the lowest call
rates, and the highest heterozygosity of any record.

## 2. Read the data and look at it

```{r}
H <- read_stacks_vcf(snps)
H
pops <- read_popmap(popmap, H$samples)
```

**What.** `read_stacks_vcf()` turns the VCF into allele matrices
(`H$A1`, `H$A2`: one row per record, one column per individual) and
works out which RAD locus each record belongs to. `read_popmap()`
matches the popmap to the VCF's sample names.

**Reading the output.** Check three things before going on:

- the number of samples and populations is what you expect (a sample
  missing from the popmap is excluded and reported);
- "records on RAD loci": here `r format(nrow(H$A1), big.mark = ",")`
  SNPs sit on `r length(unique(H$locus_raw))` loci. If a SNP VCF reads
  1.00 records per locus for RAD data, the locus assignment failed; see
  `locus_from` in `?read_stacks_vcf`. A haplotype VCF always reads 1.00
  (one record per locus), and a de novo one from Stacks says the loci
  came from "CHROM + POS", because Stacks leaves its ID column empty;
  both are expected (see "What Stacks writes" in `?read_stacks_vcf`);
- the missing genotype rate
  (`r sprintf("%.1f%%", 100 * mean(is.na(H$A1)))` here).

Every analysis function accepts either a file path or this object, and a
popmap path or this list.

## 3. Optional: filter records and genotypes

**What.** Filters take the data object and return a filtered one, so
they chain. They use **every sample in the VCF**, including any left out
of the popmap (an outgroup or a failed library, say), whereas the
analyses use only the popmap's individuals. If the VCF holds such
samples, remove them first with `H <- filter_samples(H, pops)`, so that
allele frequencies and call rates and any file written with a
`write_*()` function, describe only the individuals you analyze. Look
before filtering:

```{r}
allele_stats <- locus_allele_stats(H)
summary(allele_stats$maf)
H_f <- filter_genotype_depth(H, min_dp = 6)
H_f <- filter_call_rate(H_f, min_call = 0.8, popmap = pops)
H_f
```

Printing the data lists every filter applied since reading, with what
each one removed, which is what a methods section would need.

**Why these two.** `filter_genotype_depth()` is the same rule as Stacks'
`--min-gt-depth`, for VCFs made without it. It masks a call whatever its
genotype, and the message confirms that heterozygous and homozygous
calls were masked at similar rates (a filter that removed heterozygotes
more often would bias Ho and FIS). `filter_call_rate(..., popmap = )`
applies the missing-data threshold within each population, as `-r` does.

**Filters that change diversity statistics.** `filter_low_conf_alt()`
masks only calls that carry an ALT allele, which removes heterozygotes
preferentially. This filter allows you to remove alleles that were
calculated on a small number of reads, which may be a result of
sequencing error. This will change heterozygosity calculations, but it
may be warranted if there are concerns about genotyping errors.
`filter_maf()` and `filter_mac()` are standard, but they remove rare
alleles, which lowers per-site He, allelic richness and private allelic
richness (see the table above). If you use any of these filters, report
the threshold.

## 4. Look for close relatives

**Why.** Every test below treats individuals as independent samples of
their population. A pair of siblings is not two independent samples, and
one family can make a population look less diverse or more inbred than
it is.

```{r}
kin <- kinship_check(snps, pops)
kin$flagged_pairs
```

**Reading the output.** `kinship` is on KING's scale: about 0.5 for
duplicate samples, 0.25 for parent-offspring or full siblings, 0.125 for
half-siblings, and 0 for unrelated individuals; pairs above 0.0442
(third degree or closer) are listed. Here the planted full siblings,
`north_01` and `north_02`, read
`r round(kin$flagged_pairs$kinship[1], 2)`. With real data, keep one
individual of each flagged pair for the analyses below, or show that
results do not change without them.

**Caution.** KING misses many true relatives in RAD-seq data (McMaster
et al. 2025), so an empty `flagged_pairs` is reassuring, not proof. Use
the SNP VCF: KING needs biallelic records. With `method = "beta"`,
always pass the popmap.

## 5. Diversity within each population

`diversity_stats()` is run twice, because the two VCFs are good at
different things. The haplotype VCF gives FIS and allelic richness (a
RAD locus with several haplotypes carries more information than a
biallelic SNP); the SNP VCF gives Ho, He and their per-site values (a
per-SNP He means the same thing in every study).

### Choosing g

**What.** `g` is the number of gene copies every population is rarefied
to. It can be at most twice the size of the smallest population, but a
record only counts for a population with at least `g` gene copies there,
and for private alleles *every* population needs `g`. Missing data
therefore make a large `g` expensive. Compare two values:

```{r}
for (g in c(10, 20)) {
  res <- diversity_stats(haps, pops, g = g, nboot = 0, verbose = FALSE)
  print(res$richness[, c("population", "Ar", "Ar_n", "privAr", "privAr_n")])
}
```

**Reading the output.** `Ar_n` and `privAr_n` count the loci each value
rests on. At `g = 20` private richness rests on far fewer loci than at
`g = 10`, because only loci with at least 10 typed individuals in *both*
populations count. Pick the largest `g` that keeps most loci (here 16,
eight individuals' worth), and report it.

**Recommendation.** Use 2(r\*p) where is p is the number of populations
in the popmap, corresponding to the -p parameter passed to *populations*
if following above recommendations, and where r is the percentage of
individuals required to have a locus in a population to pass filters,
corresponding to the -r value passed to *populations*. The product is
multiplied by 2 because individuals are assumed to be diploid. Because
of the order of filtering done by *populations*, a small number of loci
will be present in less than r\*p, which is why automatic g calculations
by functions in other packages can be problematic. This is also why `g`
does not have default value. `g` must be at least 2.

### The haplotype run

```{r}
div_haps <- diversity_stats(haps, pops, g = 16, nboot = 200)
div_haps
```

**Reading the output.**

- `Fis` is `r round(div_haps$per_population$Fis[1], 3)` in north and
  `r round(div_haps$per_population$Fis[2], 3)` in south. `_se` is a
  jackknife standard error over RAD loci; `_lo` and `_hi` are a 95%
  bootstrap interval over RAD loci. (Use `nboot = 10000`, the default,
  for real analyses; 200 keeps this vignette fast.)
- `Ar` and `privAr` are per locus. North, the larger and
  non-bottlenecked population, has more alleles and more private
  alleles.
- The printed line at the bottom says which numbers to take from this
  file.
- These intervals describe one population's estimate. They are **not** a
  test of whether the populations differ; that is section 7.

`summary(div_haps)` prints the full report, with a note under each
table. One table there checks for potential null alleles:

```{r}
div_haps$fis_by_call_rate
```

Each population's loci are grouped by how many of its individuals were
genotyped there. A null allele (for example a mutation in the
restriction site) makes heterozygotes look homozygous or go missing, so
it shows up as FIS **rising** as call rate falls. Real inbreeding raises
FIS at every locus alike. Here FIS does not rise steadily as call rate
falls, so there is no sign of null alleles (the groups with few loci
have wide standard errors).

### The SNP run, and values per sequenced site

```{r}
div_snps <- diversity_stats(snps, pops, g = 16, nboot = 200, sites = sumstats,
                            verbose = FALSE)
div_snps$per_population[, c("population", "Ho", "He", "pct_poly")]
div_snps$autosomal
```

**What.** He averaged over SNPs depends on how many SNPs the run called,
so it cannot be compared with other studies. Dividing by the number of
sequenced sites, monomorphic ones included, gives heterozygosity per
sequenced site (`He_autosomal`), the quantity Schmidt et al. (2021)
recommend comparing. `sites = sumstats` reads each population's own
count of sequenced sites from Stacks'
`populations.sumstats_summary.tsv`.

The package adds up He over the SNPs in your data and divides by
`sites`. Stacks' count of sites is appropriate here only if nothing was
filtered after `populations`. Otherwise give `sites` yourself: the
sequenced sites of the loci you kept, counting a site for a population
only where at least one of its individuals is genotyped.
`diversity_stats()` warns in both cases.

**Reading the output.** South has
`r sprintf("%.0f%%", 100 * (1 - div_snps$autosomal$He_autosomal[2] / div_snps$autosomal$He_autosomal[1]))`
lower per-site He than north. `pct_poly` (percentage of SNPs polymorphic
within the population) is lower in south too: after a bottleneck, many
variants are lost altogether.

## 6. Which standard errors to report

**Why.** A standard error (SE) says how much a number would change if
you repeated the study. A population's He is uncertain for two reasons:
you typed some of the genome's RAD loci, and you caught some of the
population's individuals. The `_se`, `_lo` and `_hi` columns above
resample loci and hold the individuals fixed, so they cover only the
first reason.

That matters most when individuals differ in inbreeding. An inbred
individual is homozygous at many loci at once, so its loci are not
independent, and adding loci does not remove the uncertainty it brings.
The statistic g2 measures this; it is 0 when individuals do not differ.

```{r}
id <- identity_disequilibrium(haps, pops, nboot = 1000, nperm = 1000)
id
```

**Reading the output.** `g2` is the estimate, `g2_lo` and `g2_hi` a 95%
bootstrap interval over individuals, and `p_value` a one-sided
permutation test of g2 = 0. An interval that lies above 0 means
individuals differ in inbreeding; here that holds for
`r paste(id$population[id$g2_lo > 0], collapse = " and ")`.

**What to report.** If individuals differ in inbreeding, turn on
`se_individuals = TRUE`. It adds `_se_ind`, a jackknife over individuals
(each one left out in turn), and `_se_combined`, which puts both sources
together. For Ho, He and FIS, report `_se_combined`. For Ar and privAr,
report the locus-based `_se` (or `_lo` and `_hi`), which already covered
the truth well in simulation.

```{r}
div_ind <- diversity_stats(haps, pops, g = 16, nboot = 0, se_individuals = TRUE,
                           verbose = FALSE)
div_ind$per_population[, c("population", "Fis", "Fis_se", "Fis_se_ind", "Fis_se_combined")]
```

Here `Fis_se_combined` is about
`r round(min(div_ind$per_population$Fis_se_combined / div_ind$per_population$Fis_se), 1)`
to
`r round(max(div_ind$per_population$Fis_se_combined / div_ind$per_population$Fis_se), 1)`
times `Fis_se`: most of the uncertainty in FIS comes from which
individuals were sampled. How each SE is calculated, and the simulations
behind this rule, are in `vignette("rationale")`, section 4.

## 7. Do the populations differ?

***This test has not been as thoroughly as I'd like. Use with caution
and consider if 1) this test is necessary for your study and 2) if the
approach is sound. –N. Whelan (September 18, 2026)***

**Why individuals and loci.** The usual approaches (a bootstrap over
loci, or a paired test on per-locus values) treat thousands of loci as
thousands of replicates. But every locus is measured on the same few
individuals. When individuals differ in inbreeding, such tests find
"significant" differences between populations with identical diversity
about half the time. The opposite shortcut, a *t*-test on individuals
alone, treats the loci as fixed, and goes wrong when populations are
differentiated (`het_between_pops_selftest()` shows both).
`het_between_pops()` gives each individual one number, its
heterozygosity, and its test counts both which individuals and which
loci were sampled (`vignette("rationale")`, section 5).

```{r}
het <- het_between_pops(snps, pops, verbose = FALSE)
het
```

**Reading the output.**

- `population_summary`: mean individual heterozygosity and its spread
  (`sd`) among individuals.
- `pairwise_tests`: `diff` is north minus south, with its standard error
  (`se_combined`), 95% confidence interval (`ci_lo`, `ci_hi`) and
  p-value (`p_combined`). Report these. `p_welch` (Welch's *t*) and
  `p_wilcox` (Wilcoxon rank-sum) use individuals only and are there for
  comparison. Hedges' *g* is the difference in standard-deviation units
  (about 0.2 small, 0.5 medium, 0.8 large). With more than two
  populations, `p_combined_BH` corrects for the number of pairs.
- `pairwise_F_tests`: the same test on each individual's inbreeding
  coefficient F. Heterozygosity (observed, so He × (1 − F)) asks whether
  populations differ in heterozygosity, which reflects diversity and
  inbreeding together; F asks whether they differ in inbreeding.
- With **three or more populations**, an `omnibus` table appears first:
  one overall test (Welch's one-way ANOVA and Kruskal-Wallis) of whether
  any population differs. Interpret the pairwise tests only if it is
  significant. It uses individuals only, so confirm a difference with
  `p_combined_BH`.

`summary(het)` adds two checks worth reading before reporting the test:

```{r}
het$missingness_confound$overall
het$g2
```

- **Missingness confound.** If individuals with more missing data are
  also less heterozygous (a positive `r`), poor libraries, not biology,
  may be driving the difference. Here `r` is
  `r round(het$missingness_confound$overall$r, 2)`, not positive.
- **g2.** Whether individuals differ in inbreeding (section 6). When
  they do (`g2_lo` above 0), an inbred individual is homozygous at many
  loci at once, so loci are not independent replicates and a locus-based
  comparison would find differences too often. That is why this test
  uses individuals.

## 8. Differentiation between populations

```{r}
dif <- differentiation_stats(haps, pops, nboot = 200, verbose = FALSE)
dif$global
```

**What.** FST (Weir & Cockerham 1984) is the share of genetic variance
found between populations. Jost's D (2008) measures differentiation on a
scale that does not shrink as loci become more variable, which matters
for many-haplotype RAD loci. With the hierfstat package installed, Weir
& Goudet's (2017) beta is added to `dif$pairwise`. The global `FIS` here
is Weir & Cockerham's, pooled over populations.

**Reading the output.** FST of `r round(dif$global$FST, 3)` is modest
differentiation. On haplotype data report D (or beta) alongside FST. On
large datasets, `beta = FALSE` skips the slow hierfstat step.

**Which records D uses.** The global D uses **only records genotyped in
every population**; `D_records` in `dif$global` says how many (here
`r dif$global$D_records` of `r dif$settings$n_records`), and printing
the result says it too. A record missing some populations measures
differentiation among a different set of populations, so it is left out
of the global D. Each pairwise D in `dif$pairwise` uses every record
typed in both populations of its pair. With three or more populations
and a Stacks run with `-p` below their number, the global D can rest on
a small share of the records; report pairwise D as well. FST uses every
record typed in at least two populations (`?differentiation_stats`).

## 9. Nucleotide diversity from an all-sites VCF

**Why.** Section 5 turned He per SNP into a per-site value by dividing
by the number of sequenced sites. That cannot account for missing
genotypes at monomorphic sites. Stacks' `--vcf-all` writes every
sequenced site, and `pi_allsites()` then computes nucleotide diversity
(π) site by site, handling missing data at each one (the approach of
pixy; Korunes & Samuk 2021). Note that the `--vcf-all` output from
*populations* can create very large files.

```{r}
pi_res <- pi_allsites(allsites, pops, nboot = 200, verbose = FALSE)
pi_res$pi[, c("population", "pi", "pi_lo", "pi_hi", "pi_nc", "sites")]
pi_res$dxy[, c("pop1", "pop2", "dxy", "da", "da_nc")]
```

**Reading the output.** `pi` is the estimator of pixy, VCFtools and
Stacks; `pi_nc` leaves out the comparison between the two gene copies
inside each individual, which keeps it unbiased when individuals are
inbred. Here `pi_nc` for north (`r signif(pi_res$pi$pi_nc[1], 3)`) is
close to `He_autosomal` from section 5
(`r signif(div_snps$autosomal$He_autosomal[1], 3)`), as it should be
with little missing data. `dxy` is the divergence between populations;
`da` subtracts the within-population diversity (`pi`), and `da_nc`
subtracts `pi_nc` instead, so it is not pushed up by inbreeding.

## 10. Hardy-Weinberg: a report, never a filter

```{r}
hw <- hwe_test(haps, pops, n_draws = 1000, verbose = FALSE)
table(hw$population, significant = hw$p_value < 0.05)
```

**What.** An exact test of Hardy-Weinberg proportions for every locus in
each population, with `F` giving the direction of any departure
(positive: heterozygote deficit; negative: excess).

**Why it is not used to remove loci.** A heterozygote deficit is what
inbreeding produces, so removing deficit loci removes the signal FIS
measures. Loci with a strong heterozygote *excess* in every population
are more likely paralogs; look at them, and at `filter_max_het()`.
Testing all samples pooled (no popmap) is worse still: allele-frequency
differences between populations look like heterozygote deficits (the
Wahlund effect; Pearman et al. 2022).

## 11. Export

```{r, eval = FALSE}
write_vcf(H_f, "filtered.vcf")
write_plink(H_f, "filtered", popmap = pops)       # PLINK: add --allow-extra-chr
write_structure(H_f, "filtered.str", popmap = pops)
write_genepop(H_f, "filtered.gen", popmap = pops)
write_fstat(H_f, "filtered.dat", popmap = pops)
write_radpainter(read_stacks_vcf(haps, verbose = FALSE), "filtered.radpainter")
```

`write_plink()` stops on records with more than two alleles unless
`drop_multiallelic = TRUE`. For STRUCTURE, set `MARKERNAMES=1`,
`LABEL=1` and `POPDATA=1` in `mainparams`.

## 12. Large datasets

`read_stacks_vcf()` reads the file in chunks and keeps genotypes as
integer matrices, about 20 bytes per genotype call: 100,000 SNPs in 200
individuals take about 400 MB once read. Read a VCF once and pass the
object to every function, rather than the path each time.
`differentiation_stats(beta = FALSE)` skips its slowest step.

## A reporting checklist

- The Stacks version and `populations` settings, and any filters applied
  before or after (with thresholds).
- Which VCF each statistic came from (haplotype or SNP).
- The He estimator (Nei & Chesser 1983) and that FIS is a ratio of sums
  over loci.
- `g` for allelic richness, and `Ar_n`/`privAr_n`.
- The `sites` value(s) behind per-site heterozygosity.
- Whether relatives were removed, the g2 result, and which standard
  errors are reported (over loci or individuals).
- For population comparisons: the test on individuals, its confidence
  interval and effect size, and the missingness check.

## References

Gautier, M. et al. (2013) The effect of RAD allele dropout on the
estimation of genetic variation within and between populations.
*Molecular Ecology* 22:3165–3178.

Jost, L. (2008) G<sub>ST</sub> and its relatives do not measure
differentiation. *Molecular Ecology* 17:4015–4026.

Korunes, K.L. & Samuk, K. (2021) pixy: unbiased estimation of nucleotide
diversity and divergence in the presence of missing data. *Molecular
Ecology Resources* 21:1359–1368.

McMaster, E.S. et al. (2025) Evaluating kinship estimation methods for
reduced-representation SNP data in non-model species. *Molecular Ecology
Resources*. <doi:10.1111/1755-0998.70038>

Nei, M. & Chesser, R.K. (1983) Estimation of fixation indices and gene
diversities. *Annals of Human Genetics* 47:253–259.

Pearman, W.S., Urban, L. & Alexander, A. (2022) Commonly used
Hardy–Weinberg equilibrium filtering schemes impact population structure
inferences using RADseq data. *Molecular Ecology Resources*
22:2599–2613.

Rochette, N.C., Rivera-Colón, A.G. & Catchen, J.M. (2019) Stacks 2:
analytical methods for paired-end sequencing improve RADseq-based
population genomics. *Molecular Ecology* 28:4737–4754.

Schmidt, T.L., Jasper, M.-E., Weeks, A.R. & Hoffmann, A.A. (2021)
Unbiased population heterozygosity estimates from genome-wide sequence
data. *Methods in Ecology and Evolution* 12:1888–1898.

Weir, B.S. & Cockerham, C.C. (1984) Estimating F-statistics for the
analysis of population structure. *Evolution* 38:1358–1370.

Weir, B.S. & Goudet, J. (2017) A unified characterization of population
structure and relatedness. *Genetics* 206:2085–2103.
